---
title: "Feature Engineering: Foreign Tourists’ Nights Spent"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

Feature Engineering Description: Foreign Tourists' Nights Spent (D8R2)

The D8R2 feature captures the number of nights spent by foreign tourists
at each establishment (REGON) and region of origin (KKR). Feature
engineering was applied to extract temporal and spatial patterns from
this data to enable more detailed analysis.

**Steps performed:**

1.  Time variable extraction: The original period variable (OKRES) was
    split into year and month, and a proper date column was created
    (first day of the month). A quarter variable was also generated to
    enable quarterly aggregation of tourist stays.

2.  Hierarchical aggregation: Monthly averages of nights spent were
    calculated at different administrative levels:

3.  Municipality (D8R2_mean_gmn)

4.  County (D8R2_mean_pow)

5.  Voivodeship (D8R2_mean_won) This allows analysis of tourist stay
    patterns both locally and regionally.

6.  Rolling (moving) averages: Rolling averages over the last 3, 6, 9,
    and 12 months were computed for each municipality. These features
    smooth short-term fluctuations and highlight medium- to long-term
    trends in tourist stays.

7.  Lagged features: Lagged values of D8R2 were created for each REGON
    and KKR combination to capture temporal dependencies in the number
    of nights spent.

8.  Quarterly averages: Average nights spent per quarter
    (D8R2_quarter_mean) were calculated to support seasonal and
    higher-level temporal trend analysis.

```{r}
library(readr)
data <- read_csv("synthetic data/nights.csv")

data$...1<-NULL
data$D8R2<-NULL
```

```{r}

# --- Create date and time variables ---
data <- data %>%
  mutate(
    year = as.integer(substr(OKRES, 1, 4)),
    month = as.integer(substr(OKRES, 5, 6)),
    date = make_date(year = year, month = month, day = 1),
    quarter = quarter(date, with_year = TRUE)
  )

# --- Monthly averages for each level ---
# Municipality (GMN)
gmina_avg <- data %>%
  group_by(GMN, date) %>%
  summarise(D8R2_mean_gmn = mean(D8R2, na.rm = TRUE), .groups = "drop")

# County (POW)
powiat_avg <- data %>%
  group_by(POW, date) %>%
  summarise(D8R2_mean_pow = mean(D8R2, na.rm = TRUE), .groups = "drop")

# Voivodeship (WON)
woj_avg <- data %>%
  group_by(WON, date) %>%
  summarise(D8R2_mean_won = mean(D8R2, na.rm = TRUE), .groups = "drop")

# --- Quarterly average ---
quarter_avg <- data %>%
  group_by(quarter) %>%
  summarise(D8R2_quarter_mean = mean(D8R2, na.rm = TRUE), .groups = "drop")

# --- Rolling (moving) averages for the last 3, 6, 9, and 12 months ---
rolling_avg <- data %>%
  arrange(GMN, date) %>%
  group_by(GMN) %>%
  mutate(
    mean_3m  = rollmean(D8R2, k = 3, align = "right", fill = NA),
    mean_6m  = rollmean(D8R2, k = 6, align = "right", fill = NA),
    mean_9m  = rollmean(D8R2, k = 9, align = "right", fill = NA),
    mean_12m = rollmean(D8R2, k = 12, align = "right", fill = NA)
  ) %>%
  ungroup()

# Ensure rolling_avg has exactly one row per REGON + date to avoid many-to-many join
rolling_avg_unique <- rolling_avg %>%
  select(REGON, date, mean_3m, mean_6m, mean_9m, mean_12m) %>%
  distinct(REGON, date, .keep_all = TRUE)

# Join all aggregated features to the main dataset
data_final <- data %>%
  # Join municipality-level averages
  left_join(gmina_avg, by = c("GMN", "date")) %>%
  # Join county-level averages
  left_join(powiat_avg, by = c("POW", "date")) %>%
  # Join voivodeship-level averages
  left_join(woj_avg, by = c("WON", "date")) %>%
  # Join quarterly averages
  left_join(quarter_avg, by = "quarter") %>%
  # Join rolling (moving) averages, using the cleaned unique dataset
  left_join(rolling_avg_unique, by = c("REGON", "date"))



```
